Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-10 POPL/paper.tex
Recursive type + correction of regular exp over grammar

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that different grammar formalisms \max{which ones do
    we actually show?} can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

@@

\paragraph{Recursive Types}
isorecursive types
\steven{TODO recursive types}
  % \item $\sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x)$

@@

We only take fixed points of a single variable as a
primitive operation in the type theory, but we may apply
Beki\`c's theorem \cite{Bekić1984} to define an admissible
notion of multivariate fixed point. This is particularly
useful for defining grammars that encode the states of an
automaton. In \cref{fig:multifix} we provide the
introduction and elimination principles for such a fixed
point, where $\sigma$ is the substitution that unrolls the
mutually recursive definitions one level. That is,

\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \\
  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
\end{align*}

\begin{figure}[h!]
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}

  \inferrule
  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }
  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\end{mathpar}
\caption{Multi-fixed Points}
\label{fig:multifix}
\end{figure}

The most prevalent application of these multi-fixed points
is for defining the traces through automata, as we will see
later in \cref{eq:nfatrace}.

@@

\steven{Added this transition to handle putting the categorical semantics after
  the syntax. I think it sounds weird though}

@@

That is, the characters in the alphabet $\Sigma$ really are all of the
characters. The necessity of this axiom arises when we go to take a transition,
as we must ensure that the nexts character read has a corresponding transition
--- which would be guaranteed by determinism provided there are no surprise characters.
\steven{The above paragraph sounds weird}

@@

\subsection{Equivalence Between Regular Grammars and Finite Automata}
\label{subsec:eqproofs}
In this section, we describe a version of Thompson's
construction \cite{thompsonProgrammingTechniquesRegular1968}
   where we construct an NFA that recognizes a given regular
expression. Moreover, we will show that this NFA is strongly equivalent to the
original grammar. Witnessing this construction in our syntax has two benefits
\begin{enumerate}
  \item It reinforces this high-level view that the syntax is a natural and
        general setting for formal grammar reasoning, as we demonstrate that
        this formal system subsume results from existing systems, and
  \item Following the
        development of Thompson's construction, we then need
        only establish the equivalence of NFAs and DFAs to
        complete the full regular expression parser
\end{enumerate}

@@

\begin{theorem}[Typed Thompson]
  \label{thm:typthompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ such that $g$ is isomorphic
  to $\mathsf{AccTrace}_{N}$.
\end{theorem}

\begin{proof}[Proof Sketch]
  Recall that regular grammars are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular grammars,
  we will construct an NFA that is equivalent to $g$.

@@

Above we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape. We should reflect on the generality of our syntax, as
it allows a natural encoding of Turing machines.

>>>>

  To demonstrate the usefulness of the calculus for formal grammar
  theory, we show that different grammar formalisms \max{which ones do
    we actually show?} \steven{regular expressions, $\mu$-regular expressions
    (or context free expressions), and different automata classes (finite,
    pushdown, turing)} can be expressed as linear types in
  \theoryabbv. We give simple characterizations of the regular and
  context-free grammars as expressible by restricting which
  connectives of \theoryabbv are allowed, a form of type theoretic
  characterization of these grammar classes. Further, we demonstrate
  some central proofs of formal grammar theory can be intrinsically
  verified: \emph{strong} equivalence between regular expressions and
  NFAs is expressed by a syntactic \emph{isomorphism} definable in
  \theoryabbv.

@@

\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.

\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]

@@

We only take fixed points of a single variable as a
primitive operation in the type theory, but we may apply
Beki\`c's theorem \cite{Bekić1984} to define an admissible
notion of multivariate fixed point. This is particularly
useful for defining grammars that encode the states of an
automaton, as we will see in \cref{sec:automata}. In \cref{fig:multifix} we provide the
introduction and elimination principles for such a fixed
point, where $\sigma$ is the substitution that unrolls the
mutually recursive definitions one level. That is,

\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}).X_{1} / X_{1} , \dots, \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{n} / X_{n} \}
\end{align*}

\begin{figure}
\begin{mathpar}
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k}}

  \inferrule
  {\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{j} : \simulsubst {A_{j}}{\sigma} \vdash e_{j} : B_{j}\quad \forall j
  }
  {\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\end{mathpar}
\caption{Multi-fixed Points}
\label{fig:multifix}
\end{figure}

@@

\steven{Added the above transition paragraph to handle putting the categorical semantics after
  the syntax. I think it sounds weird though}

@@

That is, the characters in the alphabet $\Sigma$ really are all of the
characters. The necessity of this axiom arises when we go to take a transition
within the DFA,
as we must ensure that the nexts character read has a corresponding transition
--- which would be guaranteed by determinism provided there are no surprise characters.
\steven{The above paragraph sounds weird}

@@

\subsection{Equivalence Between Regular Expressions and Finite Automata}
\label{subsec:eqproofs}
In this section, we describe a version of Thompson's
construction \cite{thompsonProgrammingTechniquesRegular1968}
   where we construct an NFA that recognizes a given regular
expression. Moreover, we will show that this NFA is strongly equivalent to the
original grammar. Witnessing this construction in our syntax has two benefits
\begin{enumerate}
  \item It reinforces this high-level view that the syntax is a natural and
        general setting for formal grammar reasoning, as we demonstrate that
        this formal system subsume results from existing systems, and
  \item Following the
        development of Thompson's construction, we then need
        only establish the equivalence of NFAs and DFAs to
        complete the full regular expression parser
\end{enumerate}

@@

\begin{theorem}[Typed Thompson]
  \label{thm:typthompson}
  For $g$ a regular expression $g$, there is an NFA $N$ such that $g$ is isomorphic
  to $\mathsf{AccTrace}_{N}$.
\end{theorem}

\begin{proof}[Proof Sketch]
  Recall that regular expressions are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular expressions,
  we will construct an NFA that is equivalent to $g$.

@@

Above we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape.
